\documentclass[12pt]{article}

% ---------- PACKAGES ----------
\usepackage[a4paper,margin=1in]{geometry}
\usepackage{xcolor}
\usepackage{listings}
\usepackage{tcolorbox}
\usepackage{hyperref}
\usepackage{titlesec}
\usepackage{graphicx}
\usepackage{lmodern}
\usepackage{enumitem}

% ---------- COLORS ----------
\definecolor{deepblue}{HTML}{003366}
\definecolor{lightblue}{HTML}{E6F0FF}
\definecolor{codebg}{HTML}{F7F7F9}
\definecolor{codeborder}{HTML}{C4C4C4}
\definecolor{commentgray}{HTML}{6A737D}
\definecolor{keywordblue}{HTML}{0055AA}
\definecolor{stringgreen}{HTML}{008000}

% ---------- LINK COLORS ----------
\hypersetup{
    colorlinks=true,
    linkcolor=deepblue,
    urlcolor=keywordblue,
    citecolor=keywordblue
}

% ---------- TITLE STYLE ----------
\titleformat{\section}
  {\normalfont\Large\bfseries\color{deepblue}}
  {\thesection}{1em}{}

\renewcommand{\maketitle}{
\begin{center}
  \vspace*{1cm}
  {\Huge\bfseries\color{deepblue} ECAN’s Role in Supporting Hyperon Components}\\[0.3cm]
  {\large Tutorial Team}\\[0.2cm]
  {\small October 5, 2025}\\[0.3cm]
  \rule{0.6\textwidth}{0.5pt}
  \vspace{1cm}
\end{center}
}

% ---------- METTA CODE STYLE ----------
\lstdefinelanguage{metta}{
  morekeywords={add-atom,setAv,sti-boost,importance-diffusion,getAv,cogserver-run},
  sensitive=true,
  morecomment=[l]{;;},
  morestring=[b]"
}

\lstset{
  language=metta,
  backgroundcolor=\color{codebg},
  frame=single,
  rulecolor=\color{codeborder},
  basicstyle=\small\ttfamily,
  keywordstyle=\color{keywordblue}\bfseries,
  commentstyle=\color{commentgray}\itshape,
  stringstyle=\color{stringgreen},
  numbers=none,
  breaklines=true,
  showstringspaces=false,
  framexleftmargin=8pt,
  framexrightmargin=8pt,
  framesep=6pt,
  xleftmargin=8pt,
  xrightmargin=8pt,
}
\title{How ECAN Supports Other Hyperon Components}
\order{7}
\author{ECAN Team}
\date{October 12, 2025}
% ---------- START DOCUMENT ----------
\begin{document}

\maketitle

\section{How ECAN Supports Other Hyperon Components}

The \textbf{Economic Attention Network (ECAN)} is a foundational component of the \textbf{Hyperon} cognitive architecture, enabling efficient attention-driven resource management across multiple subsystems. ECAN provides adaptive prioritization of cognitive processes, ensuring that the system focuses on what matters most.

\begin{enumerate}[label=\textbf{\arabic*.}]
  \item \textbf{Prioritizing Cognitive Resources}
    \begin{itemize}
      \item ECAN allocates computational resources such as processing power and memory toward the most relevant data or goals.
      \item This prioritization enhances efficiency, reduces resource waste, and accelerates reasoning.
    \end{itemize}
  \item \textbf{Facilitating Knowledge Processing}
    \begin{itemize}
      \item ECAN manages attention values (\textbf{STI/LTI}) in the \textbf{AtomSpace}, allowing reasoning and learning components to focus on significant knowledge atoms.
      \item This focused attention drives more effective inference and adaptive learning.
    \end{itemize}
  \item \textbf{Supporting Scalability}
    \begin{itemize}
      \item ECAN’s economic model dynamically reallocates resources as new data, tasks, or agents appear.
      \item This flexibility ensures the Hyperon system remains scalable and responsive under diverse workloads.
    \end{itemize}
  \item \textbf{Enabling Modular Integration}
    \begin{itemize}
      \item Through the \textbf{CogServer}, ECAN connects seamlessly with reasoning modules, learning agents, and perception systems.
      \item This modular integration ensures synchronized and adaptive cognitive processing.
    \end{itemize}
\end{enumerate}
\vspace{0.5cm}
\noindent\textbf{Example: Attention Prioritization in MeTTa}
\vspace{0.2cm}

ECAN acts as the cognitive economy of Hyperon — balancing attention, energy, and computational flow.
Its dynamic feedback ensures that the system continuously reallocates focus toward the most contextually
valuable atoms.


\end{document}
